---
title: "Excel Tools for QRM Decisions"
subtitle: "1. A Beginning Decision Example"
author: "Gregg Claycamp"
date: '2021-01-13'
output:
  md_document:
  pdf_document: default
slug: Excel-examples
summary: Microsoft Excel<sup>TM</sup> is perhaps the most common spreadsheet software on business desktops. Often it is the main tool for small businesses and a prototyping and team- or division-based tool for larger business. Spreadsheet software offers an easy way to control and automate repetitive processes from data collection to knowledge managemenr. Finally, Excel macros are a convenient framework for teaching basic methods in decision analyis.  
lastmod: '2021-01-13'
featured: true
image:
  caption: 
  focal_point: 
  preview_only: no
header:
  image: ''
categories:
- QRM
- MCDA
- Decision Analysis
- Excel
draft: yes
---

```{r setup, include=FALSE}
library(knitr)
  opts_chunk$set(echo = TRUE)
  opts_chunk$set(warning = FALSE)
  opts_chunk$set(tidy = TRUE)
  opts_chunk$set(message = FALSE)
library(openxlsx)
library(dplyr)
library(magrittr)
library(kableExtra)
library(tibble)
```

## Spreadsheet software is broadly accessible

Although most data scientists and statisticians might groan at the prospect of data automation and analysis using popular spreadsheet platform, such as Excel, Google Sheets and others, the popularity of spreadsheet tools is indisputable. The popularity stems in part from its

-   accessibility to non-data scientists and non-programmers;

-   availability on most platforms including Windows, iOS, Android and cloud servers;

-   portability--the same spreadsheet can be run on independently on a laptop or launched on the internet;

-   cell formulas, many of which are intuitive for basic math needs; and

-   support from a large online community of spreadsheet users, consultants and software vendors.

A full-featured application, such as Excel^TM^, can also bridge many organizational needs from simple maintenance of contacts to elaborate dashboards and fully automated data crunching (e.g., [TheSmallman.com](https://www.thesmallman.com/)). Given the wide utility of (programmable) spreadsheets, a motivated beginner in spreadsheet technology can expect to find both room for growth and ample support developing deeper expertise using such tools.

Given all the positives, why are data scientists, engineers and other "quants" so negative about spreadsheet tools? First, comparing the spreadsheet to complete statistical platforms, such as [SAS](https://www.sas.com "Statistical software - SAS"), [Tableau](https://www.tableau.com/), [MiniTab](www.minitab.com), and many others, is an apples-and-oranges exercise! Statisticians, operations research engineers and data scientists will easily dismiss Excel by comparison to sophisticated statistical packages. Although some spreadsheet software have an impressive array of built-in and add-in statistical and mathematical functions, those platforms are far from being pre-programmed for descriptive, inferential and other statistical modeling. Second, although rigorous quantitative solutions for business analysis and forecasting can be programmed in [Visual Basic for Applications](https://docs.microsoft.com/en-us/office/vba/library-reference/concepts/getting-started-with-vba-in-office), C\# or C^++^, [Javascript](javascript.com), or [Typescript](https://www.typescriptlang.org/), the compilation and deployment tools can be bewildering to the typical spreadsheet user. Third, programming modules are difficult to manage in an organization and, (albeit in a shrinking number of cases), unsafe to share across the internet. Finally, spreadsheet programs typically run in a size-limited container making them unwieldy or even useless for a "big data" project.

<iframe width="549" height="396" frameborder="0" scrolling="no" src="https://1drv.ms/x/s!AqpWBvfRUWuxhJ89IhzxRDFtFLjpGw?e=HAS2If">

</iframe>

<iframe width="549" height="396" frameboarder="0" scrolling="no" src="https://docs.google.com/spreadsheets/d/1Hpt8RfL7hNuVXEtH_yA2CJyXzw1jCfTJ2HGO9mRq6Dc/edit?usp=sharing">

</iframe>

## Comparing Spreadsheets to R and Python {style="blue"}

Perhaps the two most popular tools for data science are the open-source languages [R](https://www.r-project.org/about.html) and [Python](https://www.python.org/). R is a statistical language program that is both free and easy to learn. Python is a free programming language that has a significant set of packages for data analysis and scientific computing. Both programs can "run anywhere" and they can be used with user-friendly integrated development environments (IDEs) such as [RStudio](rstudio.com) for R and [Idle](https://docs.python.org/3/library/idle.html) for python. For example, this blog and website are developed in RStudio using the R package, [blogdown](https://bookdown.org/yihui/blogdown/).

In spite of apparent unpopularity for statistical programming and data chores, desktop spreadsheet software remains popular due to its broad accessibility to users. Not only are spreadsheet programs available on almost every desktop, laptop and tablet produced, but the documentation and training opportunities for beginning to advance users are ubiquitous. re

Others' opinion: <https://martinctc.github.io/blog/my-favourite-alternative-to-excel-dashboards/#fn1>

vs Python [Excel vs python](https://www.nobledesktop.com/learn/python/python-vs-excel)

Use python not excel :<https://www.dataquest.io/blog/excel-vs-python/>

[R versus Excel: What's the Difference](https://www.northeastern.edu/graduate/blog/r-vs-excel/)

[Understanding R programming over Excel for Data Analysis](https://www.gapintelligence.com/blog/understanding-r-programming-over-excel-for-data-analysis/)

Alternatives: DT [DataTables in the DT package](https://rstudio.github.io/DT/)

flexDashboard R package [flexDashboard](https://rmarkdown.rstudio.com/flexdashboard/)

excel not even mentioned in survey of BI tools: <https://bi-survey.com/business-intelligence-software-comparison>

## A Basic Decision Matrix

We will develop a simple decision suitable as a foundation to multi-criteria decision analysis (MCDA). Decision matrices are a basic and intuitive approach to choosing one alternative among many based the performance of the alternative across the criteria. The matrix approach is useful for a range of simple decisions from personal to corporate choices. We will begin with a common decision of choosing a new car.

For example, I might want to choose a new car based on some criteria, including *cost, safety, capacity, fuel economy* and *style*. Although there are possibly many more criteria that can be identified, an exhaustive list is not necessary for this example nor is it even helpful.[^1] The simplest way to organize the car choice example is to create a table having columns of the criteria and rows of cars (alternatives).

[^1]: Keeney (1992) discusses the need to thoroughly identify possible alternatives in an important decision making process for important and complex decisions. But, listing every possible new car as an alternative would clearly be overkill for this decision problem. AN intuitive process of applying criteria to create a shorter list is often used. For example, the cost of new cars can quickly shorten the list.

```{r table1, include=FALSE}

tbleCols <- c( "Cost", "Safety", "Capacity", "Fuel Economy", "Style")
tbl1 <- matrix(0, 6, 5)
colnames(tbl1) <- tbleCols

rownames(tbl1) <- c("Toyota", "Ford", "Nissan", "Volvo", "Subaru", "Honda")

# Check the data table
  kable(head(tbl1),  
        caption = "A Simple Decision Table/Matrix") %>%
          kable_styling(full_width = F, position = "center", font_size = 16 ) %>%
          kable_paper(bootstrap_options = c("striped", "hover", "responsive"),  full_width = F , position = "center")
```

```{r examp}
library(dygraphs)
dygraph(nhtemp, main = "New Haven Temperatures") %>% 
  dyRangeSelector(dateWindow = c("1920-01-01", "1960-01-01"))
```

<!--chapter:end:2021-Excel-Examples-1.Rmd-->

